**skater**: An R package for SNP-based Kinship Analysis, Testing, and Evaluation

### 2021-11-30 paper/paper.Rmd

---
title: "**skater**: An R package for SNP-based Kinship Analysis, Testing, and Evaluation"
author: 
  - name: Stephen D. Turner
    affiliation: Signature Science, LLC., Austin, TX 78759, USA.
  - name: V. P. Nagraj
    affiliation: Signature Science, LLC., Austin, TX 78759, USA.
  - name: Matthew Scholz
    affiliation: Signature Science, LLC., Austin, TX 78759, USA.
  - name: Shakeel Jessa
    affiliation: Signature Science, LLC., Austin, TX 78759, USA.
  - name: Carlos Acevedo
    affiliation: Signature Science, LLC., Austin, TX 78759, USA.
  - name: Jianye Ge
    affiliation: Center for Human Identification, Department of Microbiology, Immunology, and Genetics, University of North Texas Health Science Center, Fort Worth, TX 76107, USA. 
  - name: August E. Woerner
    affiliation: Center for Human Identification, Department of Microbiology, Immunology, and Genetics, University of North Texas Health Science Center, Fort Worth, TX 76107, USA. 
  - name: Bruce Budowle
    affiliation: Center for Human Identification, Department of Microbiology, Immunology, and Genetics, University of North Texas Health Science Center, Fort Worth, TX 76107, USA. 
# Abstract up to 300 words. No citations, minimal abbreviations.
abstract: | 
    | 
    | **Motivation:** SNP-based kinship analysis with genome-wide relationship estimation and IBD segment analysis methods produces results that often require further downstream processing and manipulation. A dedicated software package that consistently and intuitively implements this analysis functionality is needed.  
    | **Results:**  Here we present the skater R package for **S**NP-based **k**inship **a**nalysis, **t**esting, and **e**valuation with **R**. The skater package contains a suite of well-documented tools for importing, parsing, and analyzing pedigree data, performing relationship degree inference, benchmarking relationship degree classification, and summarizing IBD segment data.  
    | **Availability:** The skater package is implemented as an R package and is released under the MIT license at https\:\/\/github.com/signaturescience/skater. Documentation is available at https\:\/\/signaturescience.github.io/skater.
keywords: bioinformatics, kinship, R, genealogy, SNPs, single nucleotide polymorphisms, relatedness
bibliography: bibliography.bib
output: BiocWorkflowTools::f1000_article
---

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to truth. When aggregated over many relationships and inferences, this method can help benchmark performance of a particular kinship analysis method.

@@

1. A `tibble` with calculated accuracy, lower and upper bounds, the guessing rate and p-value of the accuracy vs. the guessing rate. 
2. A `tibble` with contingency table statistics calculated for each class. Details on the statistics calculated for each class can be reviewed on the help page for `?confusion_matrix`.
3. A `matrix` with the contingency table object itself.
4. A `vector` with the reciprocal RMSE (R-RMSE). The R-RMSE is calculated as `sqrt(mean((1/(Target+.5)-1/(Predicted+.5))^2)))`, and represents an alternative to classification accuracy when benchmarking relationship degree estimation. Taking the reciprocal of the target and predicted degree results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second degree) than misclassifications at more distant relationships (e.g., misclassifying a fourth-degree relative pair as fifth-degree). The +0.5 adjustment prevents division-by-zero when a 0th-degree (identical) relative pair is introduced.

To illustrate the usage, this example will start with the `kinpairs` data from above and randomly flip ~20% of the true relationship degrees.

@@

Next, running the `confusion_matrix()` function will return all four objects noted above.

@@

SDT and VPN developed the R package.

>>>>

---
title: '**skater**: An R package for SNP-based Kinship Analysis, Testing, and Evaluation'
author:
- name: Stephen D. Turner
  affiliation: Signature Science, LLC., Austin, TX 78759, USA.
- name: V. P. Nagraj
  affiliation: Signature Science, LLC., Austin, TX 78759, USA.
- name: Matthew Scholz
  affiliation: Signature Science, LLC., Austin, TX 78759, USA.
- name: Shakeel Jessa
  affiliation: Signature Science, LLC., Austin, TX 78759, USA.
- name: Carlos Acevedo
  affiliation: Signature Science, LLC., Austin, TX 78759, USA.
- name: Jianye Ge
  affiliation: Center for Human Identification, Department of Microbiology, Immunology,
    and Genetics, University of North Texas Health Science Center, Fort Worth, TX
    76107, USA.
- name: August E. Woerner
  affiliation: Center for Human Identification, Department of Microbiology, Immunology,
    and Genetics, University of North Texas Health Science Center, Fort Worth, TX
    76107, USA.
- name: Bruce Budowle
  affiliation: Center for Human Identification, Department of Microbiology, Immunology,
    and Genetics, University of North Texas Health Science Center, Fort Worth, TX
    76107, USA.
abstract: "| \n| **Motivation:** SNP-based kinship analysis with genome-wide relationship
  estimation and IBD segment analysis methods produces results that often require
  further downstream processing and manipulation. A dedicated software package that
  consistently and intuitively implements this analysis functionality is needed.  \n|
  **a**nalysis, **t**esting, and **e**valuation with **R**. The skater package contains
  a suite of well-documented tools for importing, parsing, and analyzing pedigree
  data, performing relationship degree inference, benchmarking relationship degree
  package is implemented as an R package and is released under the MIT license at
  **Results:**  Here we present the skater R package for **S**NP-based **k**inship
  classification, and summarizing IBD segment data.  \n| **Availability:** The skater
  https\\:\\/\\/github.com/signaturescience/skater. Documentation is available at
  https\\:\\/\\/signaturescience.github.io/skater.\n"
output:
  BiocWorkflowTools::f1000_article: default
bibliography: bibliography.bib
keywords: bioinformatics, kinship, R, genealogy, SNPs, single nucleotide polymorphisms,
  relatedness
---

@@

Once estimated kinship is converted to degree, it may be of interest to compare the inferred degree to truth. When aggregated over many relationships and inferences, this approach can help benchmark performance of a particular kinship analysis method.

@@

1. A `tibble` with calculated accuracy, lower and upper bounds, the guessing rate and p-value of the accuracy vs. the guessing rate. 
2. A `tibble` with contingency table statistics calculated for each class. Details on the statistics calculated for each class can be reviewed on the help page for `?confusion_matrix`.
3. A `matrix` with the contingency table object itself.
4. A `vector` with the reciprocal RMSE (R-RMSE). The R-RMSE represents an alternative to classification accuracy when benchmarking relationship degree estimation and is calculated using the formula in (1). Taking the reciprocal of the target and predicted degree results in larger penalties for more egregious misclassifications (e.g., classifying a first-degree relative pair as second degree) than misclassifications at more distant relationships (e.g., misclassifying a fourth-degree relative pair as fifth-degree). The +0.5 adjustment prevents division-by-zero when a 0th-degree (identical) relative pair is introduced.

\begin{equation}
\sqrt{\frac{\sum_{i=1}^{k}(\frac{1}{\text{Target}+0.5}-\frac{1}{\text{Predicted}+0.5})^2}{k}}
\end{equation}

To illustrate the usage, this example will start with the `kinpairs` data from above and randomly flip ~20% of the true relationship degrees:

@@

Next, running the `confusion_matrix()` function will return all four objects noted above:

@@

SDT, VPN, and MBS developed the R package.
